Categorical Foundations of E-Graphs With Binding

### 2025-07-10 POPL2026/paper/sections/introduction.tex
lots of polish

E-graphs can efficiently simultaneously represent several equivalent terms of an algebraic theory, generalising the \textit{directed acyclic graph} (DAG) representation of terms to include equivalence classes of subterms (or subgraphs).
The key concepts of e-graphs can be intuitively grasped from some simple examples: \autoref{fig:e-graph-example} shows how a series of rewrite rules can be non-destructively applied to an e-graph~\cite{EggPaper}.

@@

Using string diagrams, $\beta$-reduction is simply the re-wiring of the diagram where an application node meets an abstraction, and explicit substitution simplifies into mere composition.
The explicit occurrences of variable names in the conventional approach raise obvious questions of scope, binding and explicit substitution, whereas the use of terms inside the explicit substitution nodes also raises many technical difficulties which makes proving various mathematical properties essentially intractable.
Additionally, explicit substitutions need to be propagated recursively throughout the graph, which adds another layer of technical and conceptual overhead.
Using de Bruijn indices instead when performing $\beta$-reduction would only push the problem elsewhere, as it greatly increases the complexity of representing the reduction as a rewrite rule, and special shifting operators need to be introduced on the right-hand side of a rewrite rule, which is a meta-syntactic operation~\cite{koehler2022sketchguided}.
Recent work on \textit{slotted} e-graphs~\cite{slotted-egraphs} uses nominal techniques to address the sharing issues when representing terms that are equivalent up to renaming of variables (even free ones) by incorporating variables as first-class data of an e-graph and making binding a built-in operation.
They also solved the problem of explicit variable names, guaranteeing freshness inside binders, explicitly differentiating between bound and free variables.
An example of a slotted e-graph is given in~\autoref{fig:slotted}.

>>>>

E-graphs can efficiently simultaneously represent several equivalent terms of an algebraic theory, generalising the \emph{directed acyclic graph} (DAG) representation of terms to include equivalence classes of subterms (or subgraphs).
The key concepts of e-graphs can be intuitively grasped from some simple examples: \autoref{fig:e-graph-example} shows how a series of rewrite rules can be non-destructively applied to an e-graph~\cite{EggPaper}.

@@

Using string diagrams, $\beta$-reduction is simply the re-wiring of the diagram where an application node meets an abstraction, and explicit substitution simplifies into mere composition.
The explicit occurrences of variable names in the conventional approach raise obvious questions of scope, binding and explicit substitution, whereas the use of terms inside the explicit substitution nodes also raises many technical difficulties which makes proving various mathematical properties essentially intractable.
Additionally, explicit substitutions need to be propagated recursively throughout the graph, which adds another layer of technical and conceptual overhead.
Using de Bruijn indices instead when performing $\beta$-reduction would only push the problem elsewhere, as it greatly increases the complexity of representing the reduction as a rewrite rule, and special shifting operators need to be introduced on the right-hand side of a rewrite rule, which is a meta-syntactic operation~\cite{koehler2022sketchguided}.
Recent work on \emph{slotted} e-graphs~\cite{slotted-egraphs} uses nominal techniques to address the sharing issues when representing terms that are equivalent up to renaming of variables (even free ones) by incorporating variables as first-class data of an e-graph and making binding a built-in operation.
They also solved the problem of explicit variable names, guaranteeing freshness inside binders, explicitly differentiating between bound and free variables.
An example of a slotted e-graph is given in~\autoref{fig:slotted}.

### 2025-07-10 POPL2026/paper/sections/soundness.tex
lots of polish

In this section we explore soundness and completeness of DPOI rewriting introduced in previous section with respect to rewriting of closed $\Sigma^{+}$-terms.
First we recall the rewriting of $\Sigma$-terms modulo SMC laws.

\begin{definition}[(Closed) $\Sigma$-term rewrite rule]
A $\Sigma$-term rewrite rules is a pair $\langle l, r \rangle$ where $l$ and $r$ are $\Sigma$ terms with matching inputs and outputs (as lists of input and output types).
When $l$ and $r$ are closed $\Sigma$-terms, we call such a rewrite rule closed.
\end{definition}

\begin{definition}
	\label{def:rewrite}
	We say that a $\Sigma$-term $f$ rewrites to a $\Sigma$-term $g$ modulo SMC laws via a rewrite rule $\langle l, r \rangle$ (notation, $\leadsto_{\langle l, r \rangle}$) if they are representable as
	$
		f = c_{1};(id_{k} \otimes l);c_{2} $ and $ g = c_{1};(id_{k} \otimes r);c_{2}
	$.
\end{definition}
This notion generalises to rewriting of closed $\Sigma$-terms.

\begin{definition}
	For closed-$\Sigma$-terms $f,g$, and a closed rewrite rule $\langle l, r \rangle$, $f \leadsto_{\langle l, r \rangle} g$ modulo SMC laws if they are representable as either in Definition~\ref{def:rewrite}, or as
	$
		f = c_{1};(id_{k} \otimes d);c_{2} $ and $ g = c_{1};(id_{k} \otimes e);c_{2}
	$
	and $d \leadsto_{\langle l, r \rangle} e$
\end{definition}
\begin{lemma}
	\label{lemma:normal_form}
	Every closed $\Sigma^{+}$ term $f$ can be equivalently represented as
	$
		f_{1} + \ldots + f_{n}
	$ modulo SMC laws and $\catname{SLat}$-equations and distributivity law~\ref{law:distributivity} where none of $f_{i}$ contain the join operator.
\end{lemma}
% \begin{proof}
% Such forms are given by orienting $\catname{SLat}$-equations such that $+$ gets propagated to the top, as, for example $f \otimes (g + h) = f \otimes g + g \otimes h$.
% Most notably, we can also escape lambda abstractions by using distributivity law~\ref{law:distributivity}.
% \end{proof}
This takes us to the following definition of rewriting for closed $\Sigma^{+}$-terms.
\begin{definition}
	For closed $\Sigma^{+}$-term $f,g$, and a closed $\Sigma$-term rewrite rule $\langle l, r \rangle$ $f \leadsto_{\langle l, r \rangle}$ modulo SMC laws, $\catname{SLat}$-equations and distributivity law~\ref{law:distributivity} if they are representable as
	$
		f = f_{1} + \ldots + f_{i} + \ldots + f_{n} $ and $ g = f_{1} + \ldots + f_{j} + \ldots + f_{n}
	$
	such that there is a permutation $\sigma$ such that there exist indices $i,j$ and $\sigma(f_{i}) \leadsto f_{j}$.
\end{definition}
Note that the rewrite rule consist of closed $\Sigma$-terms and not closed $\Sigma^{+}$-terms, i.e. neither $l$ nor $r$ contain any $+$.
This is intentional as there rewrite rules are supposed to be domain-specific (like a rule $\langle 2 \otimes id;times, id \otimes 1; <\;\;< \rangle$) and not related to the free enrichment.

Next, to formulate the correspondence between rewriting systems we first define interpretation $\llbracket - \rrbracket$ of closed $\Sigma^{+}$ terms in $\WellTypedMdaEcospans$.
It follows by first defining the interpretation of operators by extending the interpretation of plain $\Sigma$ terms as morphisms in $\MdaCospans$ as given in Appendix~\ref{sec:appendix:interpretation}.
The only missing cases, $\llbracket \textsf{ev}_{A,B} \rrbracket$ and $\llbracket \Lambda_{A,B,C} \rrbracket$, are given in Fig.~\ref{fig:ev_and_lambda}.

@@

Then we can make $\WellTypedMdaEcospans$ into $\catname{SLat}$-category by defining a join of two cospans and introducing DPOI rewrite rules to satisfy the laws of $\catname{SLat}$ symmetric monoidal category.
We also add a structural rewrite schema rule corresponding to distributivity law~\ref{law:distributivity}.
The join of two cospans is defined as in Fig.~\ref{fig:f+g} and the quotienting is done through DPOI rewrite schema rules, one for each axiom.
For example, the distributivity rule for $\otimes$ is given by the following schema rule
\ifdefined\ONECOLUMN
	\[
		\adjustbox{width=0.7\linewidth}{
			\tikzfig{./figures/semilattice_rule_1_2}~.
		}
	\]
\else
	\[
		\adjustbox{width=\linewidth}{
			\tikzfig{./figures/semilattice_rule_1_2}~.
		}
	\]
\fi

We collect all such rewrite schemas into a set $\mathcal{S}$ and then a category $\WellTypedMdaEcospans / \mathcal{S}$ is an $\catname{SLat}$ SMC.
Note that it is not a \textit{closed} $\catname{SLat}$ SMC.
To turn it into a \textit{closed} $\catname{SLat}$ SMC, we would need to also quotient by the DPO rewrites that arise from the equations in Def~\ref{def:closed}.
We do not perform such quotienting as these equations bear operational meaning, \textit{e.g.}, they model $\beta$-reduction when interpreting $\lambda$-calculus.

\begin{definition}[Quotient by rewrites]
	Given a set of DPOI rewrite rules $\mathcal{E}$,  we denote by $\WellTypedMdaEcospans/\mathcal{E}$ the category $\WellTypedMdaEcospans$ quotiented by the  relation
	$
		f \sim g \quad \text{if} \quad f \Rrightarrow^{*}_{\mathcal{E}} g
	$.
\end{definition}

This explicit quotienting allows us to only consider cospans of the form $f_1 + \ldots + f_{n}$, similar to Lemma~\ref{lemma:normal_form}, such that carrier of $f_{i}$ does not contain any hierarchical edges from $\textcolor{e-color}{E}$.

@@

\begin{proposition}[Proposition 26 and Proposition 27~\cite{fscd}]
	\label{prop:fscd}
	For two closed $\Sigma$-terms $f$ and $g$ and an $\mathcal{E}$ equation $l = r$,  $f \leadsto_{\langle l, r \rangle} g$ if and only if $\llbracket f \rrbracket \Rrightarrow_{\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle} \llbracket g \rrbracket$, where the latter is defined in a subcategory of $\WellTypedMdaEcospans$ where carriers only contain edges from $\colorbox{yellow}{E} \cup \textcolor{closed-color}{E}$.
\end{proposition}

\begin{proposition}%
	\label{prop:quotient-structural}
	For two closed $\Sigma^{+}$-terms $f$ and $g$ and an $\mathcal{E}$ equation $l = r$,  $f \leadsto_{\langle l, r \rangle} g$ if and only if $\llbracket f \rrbracket \Rrightarrow_{\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle} \llbracket g \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S}$.
\end{proposition}
\begin{proof}
	By definition, we have
	$
		f = f_{1} + \ldots + f_{n} $ and $  g = f_{1} + \ldots + f_{n}$,
	such that $f_{i} \leadsto f_{j}$ for some $i, j$.
	This yields two cospans of the form $\llbracket f_{1} \rrbracket + \ldots + \llbracket f_{n} \rrbracket$ for $\llbracket f \rrbracket$ and $\llbracket g \rrbracket$ and, by noting that every $f_{i}$ is a $\Sigma$-term, and carriers in every $\llbracket f_{i} \rrbracket$ have no edges from $\textcolor{e-color}{E}$, we can apply Proposition~\ref{prop:fscd}.
\end{proof}

>>>>

In this section we explore soundness and completeness of EDPOI rewriting with respect to rewriting of closed $\Sigma^{+}$-terms.
First we recall the rewriting of $\Sigma$-terms modulo the laws of symmetric monoidal categories.

\begin{definition}[Closed $\Sigma$-term rewrite rule]
	A $\Sigma$-term rewrite rules is a pair $\langle l, r \rangle$ where $l$ and $r$ are $\Sigma$ terms with matching inputs and outputs (as lists of input and output types).
	When $l$ and $r$ are closed $\Sigma$-terms, we call such a rewrite rule closed.
\end{definition}

\begin{definition}%
	\label{def:rewrite}
	We say that a $\Sigma$-term $f$ rewrites to a $\Sigma$-term $g$ modulo the laws of symmetric monoidal categories via a rewrite rule $\langle l, r \rangle$, writing $\leadsto_{\langle l, r \rangle}$, if they are representable as
	$
		f = c_{1};(id_{k} \otimes l);c_{2} $ and $ g = c_{1};(id_{k} \otimes r);c_{2}
	$.
\end{definition}
This notion generalises to rewriting of closed $\Sigma$-terms.

\begin{definition}
	For closed-$\Sigma$-terms $f,g$, and a closed rewrite rule $\langle l, r \rangle$, $f \leadsto_{\langle l, r \rangle} g$ modulo the laws of symmetric monoidal categories if they are representable as either in~\autoref{def:rewrite}, or as
	$
		f = c_{1};(id_{k} \otimes d);c_{2} $ and $ g = c_{1};(id_{k} \otimes e);c_{2}
	$
	and $d \leadsto_{\langle l, r \rangle} e$
\end{definition}
\begin{lemma}
	\label{lemma:normal_form}
	Every closed $\Sigma^{+}$ term $f$ can be equivalently represented as
	$
		f_{1} + \ldots + f_{n}
	$ modulo the laws of symmetric monoidal $\catname{SLat}$-categories and distributivity law (\autoref{law:distributivity}) where none of $f_{i}$ contain the join operator.
\end{lemma}
% \begin{proof}
% Such forms are given by orienting $\catname{SLat}$-equations such that $+$ gets propagated to the top, as, for example $f \otimes (g + h) = f \otimes g + g \otimes h$.
% Most notably, we can also escape lambda abstractions by using the distributivity law (\autoref{law:distributivity}).
% \end{proof}
This takes us to the following definition of rewriting for closed $\Sigma^{+}$-terms.
\begin{definition}
	For closed $\Sigma^{+}$-term $f,g$, and a closed $\Sigma$-term rewrite rule $\langle l, r \rangle$ $f \leadsto_{\langle l, r \rangle}$ modulo the laws of symmetric monoidal $\catname{SLat}$-categories and distributivity law (\autoref{law:distributivity}) if they are representable as
	$
		f = f_{1} + \ldots + f_{i} + \ldots + f_{n} $ and $ g = f_{1} + \ldots + f_{j} + \ldots + f_{n}
	$
	such that there is a permutation $\sigma$ such that there exist indices $i,j$ and $\sigma(f_{i}) \leadsto f_{j}$.
\end{definition}
Note that the rewrite rule consist of closed $\Sigma$-terms and not closed $\Sigma^{+}$-terms, i.e. neither $l$ nor $r$ contain any $+$.
This is intentional as there rewrite rules are supposed to be domain-specific (like a rule $\langle 2 \otimes id;times, id \otimes 1; <\;\;< \rangle$) and not related to the free enrichment.

Next, to formulate the correspondence between rewriting systems we first define interpretation $\llbracket - \rrbracket$ of closed $\Sigma^{+}$ terms in $\WellTypedMdaEcospans$.
It follows by first defining the interpretation of operators by extending the interpretation of plain $\Sigma$ terms as morphisms in $\MdaCospans$ as given in~\autoref{sec:appendix:interpretation}.
The only missing cases, $\llbracket \textsf{ev}_{A,B} \rrbracket$ and $\llbracket \Lambda_{A,B,C} \rrbracket$, are given in~\autoref{fig:ev_and_lambda}.

@@

Then we can make $\WellTypedMdaEcospans$ into $\catname{SLat}$-category by defining a join of two cospans and introducing EDPOI rewrite rules to satisfy the laws of symmetric monoidal $\catname{SLat}$-categories.
We also add a structural rewrite schema rule corresponding to the distributivity law (\autoref{law:distributivity}).
The join of two cospans is defined as in~\autoref{fig:f+g} and the quotienting is done through EDPOI rewrite schema rules, one for each axiom.
For example, the distributivity rule for $\otimes$ is given by the following schema rule
\ifdefined\ONECOLUMN
	\[
		\adjustbox{width=0.7\linewidth}{
			\tikzfig{./figures/semilattice_rule_1_2}~.
		}
	\]
\else
	\[
		\adjustbox{width=\linewidth}{
			\tikzfig{./figures/semilattice_rule_1_2}~.
		}
	\]
\fi

We collect all such rewrite schemas into a set $\mathcal{S}$ and then determine a symmetric monoidal $\catname{SLat}$-category $\WellTypedMdaEcospans / \mathcal{S}$ by this quotienting.
Note that it is not a \emph{closed} symmetric monoidal $\catname{SLat}$-category.
To make it so, we would need to also quotient by the DPO rewrites that arise from the equations in~\autoref{def:closed}.
We do not perform such quotienting as these equations bear operational meaning, i.e.\ they model $\beta$-reduction when interpreting $\lambda$-calculus.

\begin{definition}[Quotient by rewrites]
	Given a set of EDPOI rewrite rules $\mathcal{E}$,  we denote by $\WellTypedMdaEcospans/\mathcal{E}$ the $\catname{SLat}$-category $\WellTypedMdaEcospans$ quotiented by the relation $f \sim g \quad \text{if} \quad f \Rrightarrow^{*}_{\mathcal{E}} g$.
\end{definition}

This explicit quotienting allows us to only consider cospans of the form $f_1 + \ldots + f_{n}$, similar to~\autoref{lemma:normal_form}, such that carrier of $f_{i}$ does not contain any hierarchical edges from $\textcolor{e-color}{E}$.

@@

\begin{proposition}[Proposition 26 and Proposition 27~\cite{fscd}]
	\label{prop:fscd}
	For two closed $\Sigma$-terms $f$ and $g$ and an $\mathcal{E}$ equation $l = r$,  $f \leadsto_{\langle l, r \rangle} g$ if and only if $\llbracket f \rrbracket \Rrightarrow_{\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle} \llbracket g \rrbracket$, where the latter is defined in a sub $\catname{SLat}$-category of $\WellTypedMdaEcospans$ where carriers only contain edges from $\colorbox{yellow}{E} \cup \textcolor{closed-color}{E}$.
\end{proposition}

\begin{proposition}%
	\label{prop:quotient-structural}
	For two closed $\Sigma^{+}$-terms $f$ and $g$ and an $\mathcal{E}$ equation $l = r$,  $f \leadsto_{\langle l, r \rangle} g$ if and only if $\llbracket f \rrbracket \Rrightarrow_{\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle} \llbracket g \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S}$.
\end{proposition}
\begin{proof}
	By definition, we have
	$
		f = f_{1} + \ldots + f_{n} $ and $  g = f_{1} + \ldots + f_{n}$,
	such that $f_{i} \leadsto f_{j}$ for some $i, j$.
	This yields two cospans of the form $\llbracket f_{1} \rrbracket + \ldots + \llbracket f_{n} \rrbracket$ for $\llbracket f \rrbracket$ and $\llbracket g \rrbracket$ and, by noting that every $f_{i}$ is a $\Sigma$-term, and carriers in every $\llbracket f_{i} \rrbracket$ have no edges from $\textcolor{e-color}{E}$, we can apply~\autoref{prop:fscd}.
\end{proof}
